---
title: "Shiny Part 1: Getting Started"
author: Dan Turner (dturner@u.northwestern.edu)
---

## Welcome to my tutorial on Shiny, a popular package for creating web applications that run on R code!

Because of the structure of Shiny, I found it better to show off its capabilities using R Notebooks like this one. You will see that text (like this) can be set inline with code chunks...

```{r}
# ...like this.
```

Code chunks are run by pressing the *Run* button while the curser is in that code chunk, or by the key combination *Cmd+Shift+Enter*. Each code chunk behaves kind of like an isolated `.R` script file, but the results appear beneath the code chunk, instead of in the console.

*Try it out with the code chunk below.*

```{r RUN THIS}

# Packages used in this workshop
library(shiny)     # For building interactive apps
library(tidyverse) # For tidy data wrangling, plotting, etc.
library(DT)        # For reactive tables
library(plotly)    # For reactive plots
library(shinythemes)    # For reactive plots
library(rsconnect) # For RStudio to work with shiny
library(shinyLP)   # Easy landing pages

fight_songs <- read_csv("fight-songs.csv")  # Load a dataset
plot( sort( fight_songs$sec_duration ) )    # Plot in ascending order

```

The plot above shows the length of 65 college fight songs in the US. If there was an error, you should make sure your working directory is set to the `shiny/` directory.

### About the fight song data
This data contains the fight songs from some large American universitites, including Northwestern. Fight songs are a kind of song for fans at college games. It comes from fivethirtyeight.com: [Our Guide To The Exuberant Nonsense Of College Fight Songs](https://projects.fivethirtyeight.com/college-fight-song-lyrics/) from [here](https://github.com/fivethirtyeight/data/tree/master/fight-songs)



#######################
## SIMPLE SHINY DEMO ##
#######################

## What is the relationship between the year a fight song was written, and its duration, by conference?

This data can answer this question, and we can build an app to explore it using *shiny*.

The following code chunk is densely commented to show all the steps line-by-line for how this app works, but the short version is this: we feed the user interface data from `R` in response to the user's input.

1. The user inputs their choice of conference (a group of schools that competes) with a dropdown menu, then `shiny` passes that choice to the server...
2. The server takes the user input and returns an output. In this case, the input subsets a part of the dataset, and the server returns a plot.

Scripts that take an input and return an output interactively are called 'reactive', and it involves a special trick of shiny's. Look for the curly brackets -- shiny uses these to define reactive portions of the code.

Read the following chunk and comments...

```{r Fight Songs, message=FALSE, warning=FALSE, include=FALSE}

## The first half of every shiny app is the user interface -- the ui. It's what the user sees:

# fluidPage() makes a page based on rows and columns, and it increases in width to fill the page.
# fixedPage() is its fixed-width counterpart.
ui <- fluidPage(
  
  titlePanel("Fight Songs App"),
  
  # selectInput() makes a dropdown menu and feeds the choice to the server. It needs:
  selectInput(inputId = "conf", # an inputId (which will save it as input$inputId to make things interactive)
              label = "Select a conference", # a label that prints text over the dropdown
              selected = "Big Ten", # 'selected' is optional, and is the default selection on the dropdown. Other input types require defaults, like sliderInput()
              choices = unique(fight_songs$conference)), # a vector of choies that will populate the menu

  # plotOutput displays a ggplot to the user, in this case the one saved to output$the.plot
  plotOutput("the.plot"),
  
  # Update the page (e.g. refresh the server with new input) when the user presses "Go"
  actionButton("update", "Go")
)

## The second half of every shiny app is the shiny server, which feeds data to the UI.

# server() is defined as a function that takes an input and displays an output.
server <- function(input, output) {

  # We subset the data with dplyr and refresh it when the user presses the update button, "Go"
  # Because this responds to user input, it has to be wrapped in a reactive shell, defined by curly brackets
  selectedData <- eventReactive(input$update,{ 
    fight_songs %>% 
      mutate(isNU = school == "Northwestern") %>% # Just for fun, let's create a column to make NU easier to find
      filter(conference == paste0(input$conf)) # Here is where we bring in the user input for conference
    })
  
  # Output is returned by the server -- in this case it will return a plot
  output$the.plot <- renderPlot({
    
    # Our plot will use the selectedData from above and plot years on the x axis and duration on the y as a scatter plot
    ggplot(data = selectedData(), aes(x = year,          # year on the x axis
                                      y = sec_duration,  # duration on the y axis
                                      color = isNU)) +   # special color for NU in the plot
      geom_point(aes(shape=official_song), size = 10)    # special shapes based on whether it's official
    })
  
}

shinyApp(ui = ui, server = server)
```

## To sum it up, shiny apps have three parts:
* a UI that passes user inputs to the server and displays rendered data to the user
* a server that handles the user input to render the data for the UI
* a deployment command that links the UI and server



# SHINY WORKFLOW

The general work flow for shiny creation is:

1. Build the skeleton of the UI ("What are people going to see and input?")
2. Build the skeleton of the server ("What are the building blocks of the output?")
3. Adjust the layout of the UI ("How should the inputs and outputs look?")
4. Adjust the behavior of the server ("How can I computationally optimize the reactive scripts?")

Let's talk about each piece in turn, with an emphasis on parts 1-3. 



# SHINY UI

Shiny allows users to enter many types of data, which means it is possible for us to manipulate our data in many ways. In fact, shiny's ability to manage data for the user is probably its best asset. Find a 'gallery' of input UI's here: https://shiny.rstudio.com/gallery/widget-gallery.html

*The following code chunk shows some of the options you have as a shiny programmer for harnessing user input.*

```{r UI, message=FALSE, warning=FALSE, include=FALSE}
ui <- fluidPage(
  
  # A title for our page
  titlePanel("Common input types"),
  
  # Dropdown menu
  selectInput(inputId = "conf", 
              label = "selectInput", 
              selected = "Big Ten", # This default value is not required
              choices = unique(fight_songs$conference)),
    
  # Numeric input
  numericInput(inputId = "number", 
            label = "tnumericInput",
            value = 123),

  # Slider selector
  sliderInput(inputId = "year", 
              label = "sliderInput", 
              min = 1893, 
              max = 1968, 
              value = 1905), # This default value is required
  
  # Text input
  textInput(inputId = "text", 
            label = "textInput"),

  # Check box input
  checkboxInput(inputId = "check", 
            label = "checkboxInput",
            value = TRUE),
  
  # There is also a checkboxGroupInput option, and a radioButtons option. You will see an example of this type of checkbox later.
  
  # Date input
  dateInput(inputId = "date", 
            label = "dateInput"),
  
  # There is also a dateRangeInput option. It resembles every hotel booking website you've ever used.
    
  # File input
  fileInput(inputId = "file", 
            label = "fileInput",
            accept = c(".none"))
  
  # We will talk about the output next
  )

shinyApp(ui = ui, server = server) # launch it
```




# SHINY SERVER

To display your data in the UI, two things need to happen:

1. The data has to be rendered             e.g. `renderXYZ()`
2. The rendered data has to be output      e.g. `XYZoutput()`

Here, 'output' means to display the data in whatever form it may be. This two-step process holds for text (`renderText()` --> `textOutput()`), tables (`renderTable()` -> `tableOutput()`), and other output types you will see in the next section.

Recall that the server has two arguments: input and output. User input coming into shiny from the UI is labeled `input$inputId`, where the inputId is the argument of inputId in the input function (ala numericInput(inputId = "this")). This is very similar to the system for outputs. Everything you render is assigned to `output$outputId`. 

For example, in the following expression...

`server <- function(input, output, session) {`
`  output$example <- renderText({paste(input$greet, input$name)})`
`}`

...the output of `renderText()` would be accessible as `textOutput("example")`.

In the next code chunk, you will see a variety of example output recipes.

```{r SERVER, message=FALSE, warning=FALSE, include=FALSE}

# A bare bones UI with no user input
ui <- fluidPage(
  
  HTML("<h2>Normal Plot</h2>"), # You can render custom HTML like this
  # ^^^ is the same as h2(
  # Most common HTML tags are conveniently coded as R functions for shiny
  
  # plotOutput displays a ggplot
  plotOutput("the.plot"),
  
  # Now for an interactive table
  h2("Interactive Plot(ly)"),
  plotly::plotlyOutput("the.plotly"),

  # Standard tables (static)
  h2("Normal Table"),
  tableOutput("the.table"),
  
  # Data table (interactive)
  HTML("Interactive (Data)Table"),
  DT::DTOutput("the.datatable")
  
) #/ui

server <- function(input, output) {

  # First we make the ggplot, because we will reuse it in a moment
  gg <- ggplot(data = fight_songs, aes(x = year ), stat = "count") +   # special color for NU in the plot
      geom_bar() + # this will be a histogram
      ggtitle(label = "# Fight Songs Written by year") # a title for the plot
  
  # renderPlot() renders ggplots to the UI output variable
  output$the.plot <- renderPlot({
    gg # the plot
    })
  
  # Now let's make an interactive plot using renderPlotly()
  output$the.plotly <- renderPlotly({
    
    # Coerce fight_sounds into a leaflet
    plot_ly(fight_songs, x = ~year,  type = "histogram")
    
  })
  
  # renderTable() renders a normal table to the UI output variable
  output$the.table <- renderTable({
    
    # Shiny requires it to be an xtable object
    xtable::xtable(fight_songs)
    })
  
  # Now let's do a data table
  output$the.datatable <- renderDT({
    
    # Coerce fight_sounds into a datatable
    datatable(fight_songs)
  })
  
} #/server

shinyApp(ui = ui, server = server)

```

## Practice
In Part 2, you will create a pie chart with the fight song dataset. Within each conference, what school has the longest song?



